---
title: "The Rubin Causal Model"
subtitle: "EC655"
author: "Justin Smith"
institute: "Wilfrid Laurier University"
date: "Fall 2022"
format: 
  revealjs:
    theme: [default, hygge.scss]
    smaller: true
    slide-number: true
    chalkboard:
      theme: whiteboard
      src: drawings.json
editor: visual
title-slide-attributes:
  data-background-color: "#43464B"
---

```{r , echo=FALSE, include=FALSE}

library(sf)
library(scales)
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(hrbrthemes)
library(kableExtra)
library(viridis)
library(DT)
library(plotly)
library(readxl)
library(rgeos)
library(maptools)
library(directlabels)
library(ggrepel)
library(ggthemes)
library(cancensus)
library(bookdown)
library(geomtextpath)
library(egg)
library(ggtext)
library(ragg)
library(scico)
library(haven)
library(here)
library(knitr)
library(jpeg)
library(imager)
library(Statamarkdown)
library(ggdag)
library(dagitty)
library(geomtextpath)
library(gganimate)
library(gifski)
library(vtable)

knitr::opts_chunk$set(
 echo=FALSE, 
 message = FALSE,
 warning = FALSE,
 fig.retina = 3, 
 fig.align = "center"
 )

set.seed(9810078)
```

# Introduction

## Introduction

-   The workhorse model in econometrics is linear regression

-   Two key uses of this model

    -   Prediction

    -   Modelling (causal) effect of one variable on another

-   In many empirical application the focus is causal effects

    -   The independent effect of a particular variable on the outcome

-   We will study linear regression with a focus on causality

-   First, we attempt to understand the underlying concept of causality

-   For this we use the [Rubin Causal Model]{.red}

    -   Defines causality and under what conditions we can measure it

# Model Basics

## Potential Outcomes

-   Start with a binary framework

    -   There is a "treatment" or "no treatment"

    -   An individual theoretically has an outcome with treatment and without

-   Treatment is defined generally

    -   Getting a drug

    -   Going to university

    -   Being in a large class

-   Define the following [potential outcomes]{.red}

    -   $y_{1}$ is the outcome **with treatment**

    -   $y_{0}$ is the outcome **without treatment**

    -   $w$ is a binary variable with 1 denoting treatment, and 0 no treatment

## Treatment Effects

-   We would like to know the treatment effect $y_{1} - y_{0}$ for an individual

    -   This is the .[causal effect]{.red} of the treatment

    -   Effect differs from person to person in the population

-   [Fundamental problem of causal inference]{.red}: we never observe both $y_{1}$ and $y_{0}$

-   We only observe $(y, w)$, where

    $$y = y_{0} + (y_{1} -y_{0})w$$

    -   We observe treatment status, potential outcome given that treatment status

-   The [counterfactual]{.red} outcome with opposite treatment is never observed

## Simple Differences in Average Outcomes

-   What if we naïvely compute difference in average outcomes between treated and control? $$E(y|w=1) - E(y|w=0)$$

-   Using the definition of $y$ above,

    $$E(y|w=1) - E(y|w=0)$$ $$=  E(y_{1}|w=1) - E(y_{0}|w=0)$$ $$= \left [ E(y_{1}|w=1) - E(y_{0}|w=1) \right ] + E(y_{0}|w=1) - E(y_{0}|w=0)$$

-   The first term is called the [Average Treatment Effect on the Treated (ATT)]{.red}

    -   Average effect of the treatment for those in the treatment group

## Simple Differences in Average Outcomes

-   The second term is [Selection Bias]{.red}

    -   Baseline difference between treatment and control groups

-   Simple average differences will not identify a treatment effect

    -   It is partly a treatment effect, partly differences in who gets treated

-   Ex: Comparing average incomes of university grads to high school grads

    -   Will be partly average causal effect of university

    -   Also difference in baseline earning ability without the degree

-   Lesson is that simple differences in averages do not reveal causal effects

-   Under what conditions can we measure the causal effect of $w$ on $y$?

# Randomizing Treatment Status

## Randomization and Independence of Treatment

-   A common way to isolate treatment effects is to randomize $w$

    -   Blindly put people into treatment or control group

    -   Ensures that on average the two groups are similar at baseline

-   Mathematically, potential outcomes are independent from treatment $$(y_{0}, y_{1}) \perp w$$

-   Independence means conditioning on $w$ has no effect on expectation

    $$E(y_{0}|w=1) =E(y_{0}|w=0)$$ $$E(y_{0}|w) = E(y_{0})$$ $$E(y_{1}|w) = E(y_{1})$$

## Randomization and Treatment Effects

-   With randomization, selection bias is zero

    $$E(y_{0}|w=1) - E(y_{0}|w=0) = E(y_{0}|w=1) - E(y_{0}|w=1) = 0$$

-   As a result the difference in mean $y$ is $$E(y|w=1) - E(y|w=0)$$ $$= E(y_{1}|w=1) - E(y_{0}|w=0)$$ $$= E(y_{1}) - E(y_{0})$$

-   The first term is the ATT we saw before

-   The second term is the [Average Treatment Effect (ATE)]{.red}

    -   The treatment effect across the whole population

## Recent Example in Economics Literature

-   When we randomize treatment we can measure causal effects

-   Randomization is the standard way to measure the effects of medical treatments

-   It is becoming more popular in economics

-   Ex: Bangladesh mask study (<a href="https://www.science.org/doi/10.1126/science.abi9069">Abaluck et. al., 2021</a>)

    -   Randomized promoting mask use in rural Bangladesh

    -   Compare COVID rates between treatment and control

    -   Find some positive effect of masks, especially for age 50+

-   Next we will show how to model this in a regression framework

## Link to Regression Models

-   In randomized experiments, you can get causal effects with difference in means

-   Economists typically work with regression models

-   Remember the outcome we observe $y$ is

    $$y = (y_{1} - y_{0})w + y_{0}$$

-   If we define the potential outcomes as $$y_{0} = \alpha + \eta$$ $$y_{1} = y_{0} + \rho$$

## Link to Regression Models

-   If we combine the three equations we get

    $$y = \alpha + \rho w + \eta$$

-   The difference in means for $y$ by treatment status is now

    $$E(y|w=1) - E(y|w=0)$$ $$\rho  + E(\eta |w=1) - E(\eta |w=0)$$

-   Treatment effect is the regression slope

-   Selection bias is correlation between error term and treatment

    $$E[y_{0}|w=1] -  E[y_{0}|w=0] = E[\eta|w=1] -  E[\eta|w=0]$$

## Link to Regression Models

-   Randomization means no selection bias

-   In a regression framework, this means the error is unrelated to treatment

    $$E[\eta|w=1] -  E[\eta|w=0] =0$$

-   So a regression of $y$ on $w$ would measure the causal effect

-   In reality, we cannot always randomize $w$

    -   Economists deal mostly with observational data

-   Next we look at causal effects when we cannot randomize

# Causal Effects without Randomization

## Mean Independence of Treatment

-   We cannot always randomize into treatment and control

-   Can we uncover causal effects without experiments?

-   The answer is yes, depending on assumptions

-   One possible assumption is [Mean Independence]{.red}

    $$E(y_{0}|w) = E(y_{0})$$ $$E(y_{1}|w) = E(y_{1})$$

-   Says conditional means do not depend on treatment status

    -   Weaker assumption than full statistical independence

    -   Full independence means one event has no effect on probability of another

## Mean Independence of Treatment

-   With mean independence, we get

    $$E(y|w=1) - E(y|w=0)$$ $$= \left [ E(y_{1}|w=1) - E(y_{0}|w=1) \right ] + E(y_{0}|w=1) - E(y_{0}|w=0)$$ $$= \left [ E(y_{1}|w=1) - E(y_{0}|w=1) \right ]$$ $$=  E(y_{1}) - E(y_{0})$$

-   This identifies ATT = ATE

-   Is this assumption realistic?

    -   Means both potential outcomes unrelated to treatment

    -   On average, people in treatment and control have similar treated and non-treated outcomes

    -   Whether this is realistic depends on context

## Mean Independence of $y_{0}$

-   A variation if this assumption is [mean independence of $\mathbf{y_{0}}$]{.red}

    $$E(y_{0}|w) = E(y_{0})$$

-   With this assumption

-   Meaning that $$E(y|w=1) - E(y|w=0)$$ $$= \left [ E(y_{1}|w=1) - E(y_{0}|w=1) \right ] + E(y_{0}|w=1) - E(y_{0}|w=0)$$ $$= \left [ E(y_{1}|w=1) - E(y_{0}|w=1) \right ]$$

-   With this assumption, we only measure the ATT (Not ATE)

-   Is this realistic?

    -   Means untreated outcome is same between groups on average

    -   Puts no restriction on differences in treated outcome

    -   Intuitively, there are no baseline differences between groups

## Conditional Mean Independence

-   We can also use other variables to help with our assumptions

-   Suppose we observe a set of pre-treatment characteristics $\mathbf{x}$

    -   Ex: gender, parental education, school test scores, etc.

    -   Key is they are determined before treatment

-   With this information you could assume [Conditional Independence]{.red}

    $$(y_{0}, y_{1}) \perp w |\mathbf{x}$$

-   Conditional on $\mathbf{x}$, treatment is independent of outcomes

-   Write this mathematically as $$E(y_{0}|w=1, \mathbf{x}) =E(y_{0}|w=0, \mathbf{x})$$ $$E(y_{0}|w, \mathbf{x}) = E(y_{0}| \mathbf{x})$$ $$E(y_{1}|w, \mathbf{x}) = E(y_{1}|\mathbf{x})$$

## Conditional Mean Independence

-   This implies that we can get treatment effects at each $\mathbf{x}$ $$E(y|w=1,  \mathbf{x}) - E(y|w=0,  \mathbf{x})$$ $$= E(y_{1}|w=1,  \mathbf{x}) - E(y_{0}|w=1,  \mathbf{x})= E(y_{1} | \mathbf{x}) - E(y_{0}| \mathbf{x})$$ $$= ATT( \mathbf{x}) =ATE( \mathbf{x})$$

-   These treatment effects are functions of $\mathbf{x}$

    -   They will differ across values of $\mathbf{x}$

    -   So there are multiple treatment effects

-   Finally, a variation on this is [Conditional Mean Independence]{.red}

    $$E(y_{0}|w, \mathbf{x}) = E(y_{0}| \mathbf{x})$$ $$E(y_{1}|w, \mathbf{x}) = E(y_{1}|\mathbf{x})$$

-   Gives you the same $ATE( \mathbf{x}) = ATT( \mathbf{x})$ as above

## Link with Regression

-   We can also model this scenario with regression

-   Remember again the following equations for $y$, $y_{0}$, and $y_{1}$

    $$y = (y_{1} - y_{0})w + y_{0}$$ $$y_{0} = \alpha + \eta$$ $$y_{1} = y_{0} + \rho$$

-   This time, the random term $\eta$ depends on $\mathbf{x}$

    $$\eta = \mathbf{x}\boldsymbol{\gamma} + \epsilon$$

## Link with Regression

-   Substitute this into the equation for $y_{0}$

    $$y_{0} = \alpha + \mathbf{x}\boldsymbol{\gamma} + \epsilon$$

-   Now if we combine the equations for $y$, $y_{0}$, and $y_{1}$

    $$y = \alpha + \rho w + \mathbf{x}\boldsymbol{\gamma} + \epsilon$$

-   The difference in means for $y$ at each $\mathbf{x}$ between groups is now $$E(y| \mathbf{x}, w=1) - E(y| \mathbf{x}, w=0)$$ $$\rho  + E(\epsilon | \mathbf{x}, w=1)  - E(\epsilon | \mathbf{x}, w=0)$$

## Link with Regression

-   Again, the difference in the mean conditional error term is selection bias

    $$E[y_{0}|\mathbf{x}, w=1] -  E[y_{0}|\mathbf{x}, w=0] = E[\epsilon|\mathbf{x},w=1] -  E[\epsilon|\mathbf{x},w=0]$$

-   If we assume Conditional Independence (or Conditional Mean Independence) then $$E(y_{0}  | \mathbf{x}, w)  = E(y_{0}  | \mathbf{x})$$

-   and as a result

    $$E(y_{0}  | \mathbf{x}, w=1)  = E(y_{0}  | \mathbf{x}, w=0)$$

-   and

    $$E[\epsilon|\mathbf{x},w=1] -  E[\epsilon|\mathbf{x},w=0] =0$$

## Link with Regression

-   Intuition is as follows

    -   Participation in treatment and control might be related to baseline factors

    -   But, for those with the same baseline factors, treatment is as good as random

    -   Computing treatment effects for each $\mathbf{x}$ eliminates selection bias

-   In this model, adding "control variables" identifies the causal effect of $w$

-   Whether this is actually true depends on assumptions

    -   Often we cannot prove those assumptions

# Summary of Rubin Model

## Summary of Rubin Model

-   The Rubin model defines what is a causal effect

-   Roughly speaking, it an Average Treatment Effect

    -   They will differ across values of $\mathbf{x}$

    -   Difference in potential outcomes, on average in population

    -   Depending on context, it might be an Average Treatment Effect for the Treated

-   We can express the Rubin model in a regression framework

-   The slope in a linear regression is the causal effect if we can assume one of

    -   Randomization of treatment

    -   "As good as" randomization

        -   Mean Independence, Conditional Independence, Conditional Mean Indpendence

## Summary of Rubin Model

-   When our regression model identifies an underlying causal effect, we call it a [structural model]{.red}

-   In many econometric applications, this is what we want

-   Next, we discuss in more detail linear regression

-   First we discuss the population model

    -   We will define the parameters we are measuring

    -   Some of this might be new

-   Then we discuss estimation by OLS

    -   Focus is on when OLS consistently estimates the parameters

# Simulation

## Data Setup {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   To help understand the Rubin model we will demonstrate with simulated data

-   Code to the right creates potential outcomes

-   For simplicity the treatment effect is set to 5 for everyone

-   Outcomes $y_{0}$ and $y_{1}$ have a Normal distribution because of $\eta$
:::

::: {.column width="50%"}
```{r, echo=TRUE}
data <- data.frame(eta=rnorm(100000,0,1)) %>%
  mutate(y0 = 2 + eta, y1 = y0 + 5, 
         treat_eff = y1 - y0)

sumtable(data, summ=c('notNA(x)','mean(x)','sd(x)'))
```
:::
:::

## Random Assignment to Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Next assign treatment $w$ using [randomization]{.red}

-   In the code, $w=1$ randomly with probability 0.5

-   Compute observed $y$ based on treatment status
:::

::: {.column width="50%"}
```{r, echo=TRUE}
data %<>% mutate(w = if_else(runif(100000) > .5,1,0), 
                 y = y0 + (y1-y0)*w) %>% 
  group_by(w)

head(data)
```
:::
:::

## Random Assignment to Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   With random assignment we know

    -   $y_{0}$ is independent of $w$

    -   $y_{1}$ is independent of $w$

-   So the distributions of $y_{0}$ and $y_{1}$ are the same when $w=0$ and when $w=1$

-   To the right we show the distribution of $y_{0}$
:::

::: {.column width="50%"}
```{r, echo=TRUE}
ggplot(data, aes(x=y0, color=as.factor(w))) +
  geom_density(alpha = .4, size=2) +
  theme_pander(nomargin=FALSE, boxes=TRUE) +
  labs(title = "Distribution of Y0")
```
:::
:::

## Random Assignment to Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Randomization ensures difference in average $y$ between groups equals the [ATE and ATT]{.red}

-   On the right we show the difference in mean of $y$ equals 5

    -   (it's not exactly 5 due to sampling error)
:::

::: {.column width="50%"}
```{r, echo=TRUE}
summarize(data, mean(y))
summarize(data,mean(y))$`mean(y)`[2] - 
  summarize(data,mean(y))$`mean(y)`[1]
```
:::
:::

## Random Assignment to Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Can implement difference in means as a regression

-   Recall slope in OLS regression of $y$ on dummy variable is difference in means of $y$
:::

::: {.column width="50%"}
```{r, echo=TRUE}
lm(y ~ w, data)
```
:::
:::

## Selection into Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Now simulate selection into treatment based on $y_{0}$

    -   Treatment now related to value of $y_{0}$

-   We know $\eta$ determines value of $y_{0}$

    -   If we make $w=$ with higher values of $\eta$ then $w$ is related to $y_{0}$
:::

::: {.column width="50%"}
```{r, echo=TRUE, results=FALSE}
data2 <- data %>% 
  ungroup() %>% 
  select(eta, y0,y1) %>%
  mutate(w = if_else(eta + runif(100000,-1,1) > 0,1,0), 
         y = y0 + (y1-y0)*w) %>%
  group_by(w)

sumtable(data2, 
         summ=c('notNA(x)','mean(x)','sd(x)'), 
         group="w",
         group.long = TRUE)
```
:::
:::

## Selection into Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   The means of $y_{0}$ and $y_{1}$ are now different by group

-   Because of selection bias

    -   Treated group has better non-treated outcomes
:::

::: {.column width="50%"}
```{r, results=TRUE}
sumtable(data2, 
         summ=c('notNA(x)','mean(x)','sd(x)'), 
         group="w",
         group.long = TRUE)
```
:::
:::

## Selection into Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   The distribution of $y_{0}$ differs by $w$

    -   Treated group has better baseline outcomes

-   This creates selection bias
:::

::: {.column width="50%"}
```{r, echo=TRUE}
ggplot(data2, aes(x=y0, color=as.factor(w))) +
  geom_density(alpha = .4, size=2) +
  theme_pander(nomargin=FALSE, boxes=TRUE) +
  labs(title = "Distribution of Y0")
```
:::
:::

## Selection into Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Selction bias shows up when you take difference in mean $y$

    -   We know the true treatment effect is 5

    -   But difference in $y$ is larger

    -   There is positive selection bias

    -   Bias is about 1.4
:::

::: {.column width="50%"}
```{r, echo=TRUE}
summarize(data2, mean(y))
summarize(data2,mean(y))$`mean(y)`[2] - 
  summarize(data2,mean(y))$`mean(y)`[1]
```
:::
:::

## Selection into Treatment {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   You can implement this as a regression

-   OLS estimates biased treatment effect

-   Remember the intercept is mean of $y$ when $w=0$
:::

::: {.column width="50%"}
```{r, echo=TRUE}
lm(y ~ w, data2)
```
:::
:::

## Mean Independence of $y_{0}$ {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Randomization ensures entire distribution of $y_{0}$ is the same for treatment and control

-   We do not need this to estimate **average** treatment effects

-   If the **mean** of $y_{0}$ is the same between treatment and control we can estimate treatment effect with difference in means

-   Code makes mean of $y_{0}$ the same, but variance bigger for $w=1$
:::

::: {.column width="50%"}
```{r, echo=TRUE, results=FALSE}
data3 <- data %>% 
  ungroup() %>% 
  select(eta, y0,y1) %>%
  mutate(w =  if_else(between(percent_rank(y0),.25,.75),0,1), 
         y = y0 + (y1-y0)*w) %>%
  group_by(w)

sumtable(data3, 
         summ=c('notNA(x)','mean(x)','sd(x)'), 
         group="w",
         group.long = TRUE)
```
:::
:::

## Mean Independence of $y_{0}$ {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Randomization ensures entire distribution of $y_{0}$ is the same for treatment and control

-   We do not need this to estimate **average** treatment effects

-   If the **mean** of $y_{0}$ is the same between treatment and control we can estimate treatment effect with difference in means

-   Code makes mean of $y_{0}$ the same, but variance bigger for $w=1$
:::

::: {.column width="50%"}
```{r, results=TRUE}
sumtable(data3, 
         summ=c('notNA(x)','mean(x)','sd(x)'), 
         group="w",
         group.long = TRUE)
```
:::
:::

## Mean Independence of $y_{0}$ {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   The distribution of $y_{0}$ is plotted on the right

-   The spread is larger for $w=1$

-   This does not affect estimate of the **average** treatment effect
:::

::: {.column width="50%"}
```{r, echo=TRUE, fig.height = 4.55}
ggplot(data3, aes(x=y0, color=as.factor(w))) +
  geom_density(alpha = .4, size=2) +
  theme_pander(nomargin=FALSE, boxes=TRUE) +
  labs(title = "Distribution of Y0")
```
:::
:::

## Mean Independence of $y_{0}$ {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Take difference in mean $y$

    -   This equals the treatment effect

    -   Difference in variance did not create bias
:::

::: {.column width="50%"}
```{r, echo=TRUE}
summarize(data2, mean(y))
summarize(data2,mean(y))$`mean(y)`[2] - 
  summarize(data2,mean(y))$`mean(y)`[1]
```
:::
:::

## Mean Independence of $y_{0}$ {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Running regression produces same result

-   The variance in $y_{1}$ affects the standard error

    -   But we are not concerned with that right now
:::

::: {.column width="50%"}
```{r, echo=TRUE}
lm(y ~ w, data3)
```
:::
:::

## Conditional Mean Independence {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Finally consider [conditional mean independence]{.red}

-   Treatment is related to $y_{0}$, but only through $x$

    -   For people with the same $x$, $y_{0}$ is unrelated to $w$

-   Ex: Education and wages

    -   Smart people $(x = 1)$ earn higher wages regardless of schooling $(y_0)$

    -   Smart people are more likely to go to university $(w = 1)$

    -   People at university will have higher $y_0$
:::

::: {.column width="50%"}
```{r, echo=TRUE, results=FALSE}
data4 <- data %>% 
  ungroup() %>% 
  select(eta) %>%
  mutate(x = if_else(runif(100000) > .5,1,0),
         w = if_else(x + runif(100000, -1,1) > .5,1,0),
         y0 = 2 + 3*x + eta,
         y1 = y0 + 5,
         y = y0 + (y1-y0)*w) %>%
  group_by(w)

sumtable(data4, 
         summ=c('notNA(x)','mean(x)','sd(x)'), 
         group="w",
         group.long = TRUE)
```
:::
:::

## Conditional Mean Independence {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Comparing treatment and control, $y_{0}$ is bigger when $w=1$

-   This is because

    -   $y_{0}$ is bigger when $x=1$

    -   $w$ more likely to be $1$ when $x=1$
:::

::: {.column width="50%"}
```{r, results=TRUE}
sumtable(data4, 
         summ=c('notNA(x)','mean(x)','sd(x)'), 
         group="w",
         group.long = TRUE)
```
:::
:::

## Conditional Mean Independence {background-color="#d9edc2"}

::: columns
::: {.column width="35%"}
-   What if we focus only on people with $x=1$?

-   No difference in $y_{0}$ between treated and control

    -   Because $x$ is only reason why they differed

    -   This is holding $x$ fixed
:::

::: {.column width="65%"}
```{r, results=TRUE, echo= TRUE}
sumtable(filter(data4, x==1), 
         summ=c('notNA(x)','mean(x)','sd(x)'), 
         group="w")
```
:::
:::

## Conditional Mean Independence {background-color="#d9edc2"}

::: columns
::: {.column width="35%"}
-   Same result if we hold $x=0$?

    -   Again because $x$ is only reason why they differed
:::

::: {.column width="65%"}
```{r, results=TRUE, echo= TRUE}
sumtable(filter(data4, x==0), 
         summ=c('notNA(x)','mean(x)','sd(x)'), 
         group="w")
```
:::
:::

## Conditional Mean Independence {background-color="#d9edc2"}

::: columns
::: {.column width="50%"}
-   Regression of $y$ on $w$ is biased

    -   Because $w$ is correlated with error

-   But regression of $y$ on $w$ and $x$ generates actual treatment effect

-   This is conditional mean independence

    -   Holding $x$ fixed, potential outcomes no longer related to treatment
:::

::: {.column width="50%"}
```{r, echo=TRUE}
lm(y ~ w, data4)
lm(y ~ w + x, data4)
```
:::
:::
